The Unicode Cookbook \linebreak for Linguists

### 2018-03-09 book/chapters/orthography_profiles.tex
ready for review two

Our central aims for linguistics, to be approached with a Unicode-based
solution, are: (i) to improve the consistency of the encoding of sources, (ii)
to transparently document knowledge about the writing system (including
transliteration), and (iii) to do all of that in a way that is easy and quick to
manage for many different sources with many different writing systems. The
central concept in our proposal is the \textsc{orthography profile}, a simple
tab-separated text file, that characterizes and documents a writing system.
We also offer basic implementations in Python and R to assist with the
production of such files, and to apply orthography profiles for consistency
testing, grapheme tokenization and transliteration. Not only can orthography
profiles be helpful in the daily practice of linguistics, they also succinctly
document the orthographic details of a specific source, and, as such, might
fruitfully be published alongside sources (e.g.~in digital archives). Also, in
high-level linguistic analyzes in which the graphemic detail is of central
importance (e.g.~phonotactic or comparative-historical studies), orthography
profiles can transparently document the decisions that have been taken in the
interpretation of the orthography in the sources used.

@@

In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points
(using NFD normalization), delimited here by space. Unicode tokenization is
required because sequences of code points can differ in their visual and logical
orders. For example, <õ̰> is ambiguous to whether it is the sequence of <o> +
<\dia{0303}> + <\dia{0330}> or <o> + <\dia{0330}> + <\dia{0303}>. Although these two variants are visually homoglyphs,
computationally they are different (see Sections \ref{pitfall-ipa-homoglyphs} and \ref{pitfall-homoglyphs-in-IPA} ). 
Unicode normalization should be applied to
this string to reorder the code points into a canonical order, allowing the data
to be treated canonically equivalently for search and comparison.

@@

The crucial difference between our current proposal and traditional
computational approaches to transliteration is the strict separation between
tokenization and transliteration. Most computational approaches to
transliteration are based on finite-state transducers (including the
transliteration as described in the Unicode Locale Data Markup
Language).\footnote{See \url{http://www.unicode.org/reports/tr35/}} Finite-state 
transducers attempt to describe the mapping from input to output string directly 
as a set of rewrite rules. Although such systems are computationally well 
understood, we feel that they are not well-suited for day-to-day linguistic 
practice. First, by forcing a first step of grapheme tokenization, our system 
tries to keep close to the logic of the writing system. Second, by separating 
tokenization from transliteration there is no problem with `feeding' and 
`bleeding' of rules, common with transducers (cf.~Section~\ref{r-implementation}
).

@@

In practice, we foresee a workflow in which orthography profiles are iteratively
refined, while at the same time inconsistencies and errors in the data to be
tokenized are corrected. In some more complex use-cases there might even be a
need for multiple different orthography profiles to be applied in sequence (see
Section~\ref{r-implementation} on various exemplary use-cases). The result of any such
workflow will normally be a cleaned dataset and an explicit description of the
orthographic structure in the form of an orthography profile. Subsequently, the
orthography profiles can be easily distributed in scholarly channels alongside
the cleaned data, for example in supplementary material added to journal papers
or in electronic archives.

@@

%	\item \textsc{A profile is a unicode UTF-8 encoded text file}, ideally
%       using NFC, no-BOM, and LF (see Section~\ref{pitfall-file-formats}), that
%       includes the information pertinent to the orthography. 
	\item \textsc{A profile is a CSV file with an obligatory header
       line}. A minimal profile must have a single column with the header ``Grapheme''. 
	   For any additional columns, the name in the header of the CSV file
       must be specified. The actual ordering of the columns is unimportant. Cells should 
	   not contain tabs or newlines. The header list must be delimited 
	   in the same way as the rest of the file's contents. Each record must be kept 
	   on a separate line. Separate lines with comments are not allowed. Comments that
       belong to specific lines must be put in a separate column of
       the CSV file, e.g.~add a column called \textsc{comments}.

@@

	\item \textsc{Metadata should be added in a separate UTF-8 text file} using 
	   the JSON-DL dialect specified in \textit{Metadata Vocabulary for Tabular 
	   Data}.\footnote{\url{https://www.w3.org/TR/tabular-metadata/}} This metadata 
	   format allows easy inclusion of Dublin Core metadata,\footnote{\url{http://dublincore.org/}} 
	   which should be used to specify information about the orthographic description 
	   in the orthography profile. The orthography profile metadata should minimally include: 
	   (i) author, (ii) date, (iii) title of the profile, (iv) a stable language identifier
       encoded in BCP 47/ISO 639-3 of the target language of the profile, and (v)
       bibliographic data for resource(s) that illustrate the orthography
       described in the profile. Further, (vi) the tokenization method and (vii)
       the Unicode normalization form used should be documented here (see below).
	   In accordance with the \textit{Metadata Vocabulary for Tabular 
	   Data}, the metadata's filename must consist of the orthography 
	   profile's filename appended with ``-metadata.json''.\footnote{JSON-DL metadata 
	   is also the choice for datasets conforming to the Cross-Linguistic Data Formats standard, 
	   see: \url{http://cldf.clld.org/}.}

@@

\begin{enumerate}
	\def\labelenumi{A\arabic{enumi}.} \setcounter{enumi}{4} 
	\item \textsc{A minimal profile consists of a single column} with a header
       called \texttt{Grapheme}, listing each of the different graphemes in a
       separate line. The name of this column is crucial for automatic 
       processing.
	\item \textsc{Optional columns can be used to specify the left and right
       context of the grapheme}, to be designated with the headers \texttt{Left}
       and \texttt{Right}, respectively. The same grapheme can occur multiple
       times with different contextual specifications, for example to
       distinguish different pronunciations depending on the context. 
	\item \textsc{The columns \texttt{Grapheme}, \texttt{Left} and \texttt{Right}
       can use regular expression metacharacters.} If regular expressions are
       used, then all literal usage of the special symbols, like full stops <.>
       or dollar signs <\$> (so-called \textsc{metacharacters}) have to be
       explicitly escaped by adding a backslash before them (i.e.~use
       <\textbackslash.> or <\textbackslash\$>). Note that any specification of
       context automatically expects regular expressions, so it is probably
       better to always escape all regular expression metacharacters when used
       literally in the orthography. The following symbols will need to be
       preceded by a backslash: {[} {]} ( ) \{ \} | ~+ * . - ! ? \^{} \$ and the
       backslash \textbackslash~itself. 
	\item \textsc{An optional column can be used to specify classes of graphemes},
       to be identified by the header \texttt{Class}. For example, this column
       can be used to define a class of vowels. Users can simply add ad-hoc
       identifiers in this column to indicate a group of graphemes, which can
       then be used in the description of the graphemes or the context. The
       identifiers should of course be chosen such that they do not conflate
       with any symbols used in the orthography themselves. Note that such
       classes only refer to the graphemes, not to the context. 
	\item \textsc{Columns describing transliterations for each graphemes can be
       added and named at will}. Often more than a single possible
       transliteration will be of interest. Any software application using these
       profiles should prompt the user to name any of these columns to select a
       specific transliteration. 
	\item \textsc{Any other columns can be added freely, but will mostly be ignored
       by any software application using the profiles}. As orthography profiles
       are also intended to be read and interpreted by humans, it is often
       highly useful to add extra information on the graphemes in further
       columns, like for example Unicode codepoints, Unicode names, frequency of
       occurrence, examples of occurrence, explanation of the contextual
       restrictions, or comments. 
 \end{enumerate}

@@

\begin{enumerate}
	\def\labelenumi{B\arabic{enumi}.} 
	\item \textsc{Each line of a profile will be interpreted as a regular
       expression. }Software applications using profiles can also offer to
       interpret a profile in the literal sense to avoid the necessity for the
       user to escape regular expressions metacharacters in the profile.
       However, this only is possible when no contexts or classes are described,
       so this seems only useful in the most basic orthographies. 
	\item \textsc{The \textsc{class} column will be used to produce explicit
       \textsc{or} chains of regular expressions}, which will then be inserted
       in the \texttt{Grapheme}, \texttt{Left} and \texttt{Right} columns at
       the position indicated by the class-identifiers. For example, a class
       called \texttt{V} as a context specification might be replaced by a regular
       expression like:
       \texttt{(a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au)}.
       Only the graphemes themselves are included here, not any contexts
       specified for the elements of the class. Note that in some cases the 
       ordering inside this regular expression might be crucial.
	\item \textsc{The \textsc{left} and \textsc{right} contexts will be included
       into the regular expressions by using lookbehind and lookahead}.
       Basically, the actual regular expression syntax of lookbehind and
       lookahead is simply hidden to the users by allowing them to only specify
       the contexts themselves. Internally, the contexts in the columns
       \texttt{Left} and \texttt{Right} are combined with the column
       \texttt{Grapheme} to form a complex regular expression like:\ 
       \texttt{(?\textless{}=Left)Grapheme(?=Right)}. 
	\item \textsc{The regular expressions will be applied in the order as specified
       in the profile, from top to bottom.} A software implementation can offer
       help in figuring out the optimal ordering of the regular expressions, but
       should then explicitly report on the order used.      
\end{enumerate}

@@

\begin{enumerate}
	\def\labelenumi{B\arabic{enumi}.} \setcounter{enumi}{4} 
	\item \textsc{All graphemes are matched in the text before they are tokenized
       or transliterated}. In this way, there is no necessity for the user to
       consider `feeding' and `bleeding' situations, in which the application of
       a rule either changes the text so another rule suddenly applies (feeding)
       or prevents another rule to apply (`bleeding'). 
	\item \textsc{The matching of the graphemes can occur either globally or
       linearly. }From a computer science perspective, the most natural way to
       match graphemes from a profile in some text is by walking linearly
       through the text-string from left to right, and at each position go
       through all graphemes in the profile to see which one matches, then go to
       the position at the end of the matched grapheme and start over. This is
       basically how a finite state transducer works, which is a
       well-established technique in computer science. However, from a
       linguistic point of view, our experience is that most linguists find it
       more natural to think from a global perspective. In this approach, the
       first grapheme in the profile is matched everywhere in the text-string
       first, before moving to the next grapheme in the profile. Theoretically,
       these approaches will lead to different results, though in practice of
       actual natural language orthographies they almost always lead to the same
       result. Still, we suggest that any software application using orthography
       profiles should offer both approaches (i.e. \textsc{global} or
       \textsc{linear}) to the user. The approach used should be documented in
       the metadata as \textsc{tokenization method}. 
	\item \textsc{The matching of the graphemes can occur either in NFC or NFD.} By
       default, both the profile and the text-string to be tokenized should be
       treated as NFC (see Section~\ref{pitfall-canonical-equivalence}). However, in some use-cases it turns out to
       be practical to treat both text and profile as NFD. This typically
       happens when very many different combinations of diacritics occur in the
       data. An NFD-profile can then be used to first check which individual
       diacritics are used, before turning to the more cumbersome inspection of
       all combinations. We suggest that any software application using
       orthography profiles should offer both approaches (i.e. \textsc{NFC} or
       \textsc{NFD}) to the user. The approach used should be documented in the
       metadata as \textsc{unicode normalization}. 
	\item \textsc{The text-string is always returned in tokenized form} by
       separating the matched graphemes by a user-specified symbols-string. Any
       transliteration will be returned on top of the tokenization. 
	\item \textsc{Leftover characters}, i.e.~\textsc{characters that are not
       matched by the profile, should be reported to the user as errors.}
       Typically, the unmatched character are replaced in the tokenization by a
       user-specified symbol-string.       
 \end{enumerate}

>>>>

Our central aims for linguistics, to be approached with a Unicode-based
solution, are: (i) to improve the consistency of the encoding of sources, (ii)
to transparently document knowledge about the writing system (including
transliteration), and (iii) to do all of that in a way that is easy and quick to
manage for many different sources with many different writing systems. The
central concept in our proposal is the \textsc{orthography profile}, a simple
delimited text file, that characterizes and documents a writing system.
We also offer basic implementations in Python and R to assist with the
production of such files, and to apply orthography profiles for consistency
testing, grapheme tokenization and transliteration. Not only can orthography
profiles be helpful in the daily practice of linguistics, they also succinctly
document the orthographic details of a specific source, and, as such, might
fruitfully be published alongside sources (e.g.~in digital archives). Also, in
high-level linguistic analyzes in which the graphemic detail is of central
importance (e.g.~phonotactic or comparative-historical studies), orthography
profiles can transparently document the decisions that have been taken in the
interpretation of the orthography in the sources used.

@@

In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points
(using NFD normalization), delimited here by space. Unicode tokenization is
required because sequences of code points can differ in their visual and logical
orders. For example, <õ̰> is ambiguous to whether it is the sequence of <o> +
<\dia{0303}> + <\dia{0330}> or <o> + <\dia{0330}> + <\dia{0303}>. Although these two variants are visually homoglyphs,
computationally they are different (see Sections \ref{pitfall-ipa-homoglyphs} and \ref{pitfall-homoglyphs-in-IPA}). 
Unicode normalization should be applied to
this string to reorder the code points into a canonical order, allowing the data
to be treated canonically equivalently for search and comparison.

@@

The crucial difference between our current proposal and traditional
computational approaches to transliteration is the strict separation between
tokenization and transliteration. Most computational approaches to
transliteration are based on finite-state transducers (including the
transliteration as described in the Unicode Locale Data Markup
Language).\footnote{See \url{http://www.unicode.org/reports/tr35/}} Finite-state 
transducers attempt to describe the mapping from input to output string directly 
as a set of rewrite rules. Although such systems are computationally well 
understood, we feel that they are not well-suited for day-to-day linguistic 
practice. First, by forcing a first step of grapheme tokenization, our system 
tries to keep close to the logic of the writing system. Second, by separating 
tokenization from transliteration there is no problem with `feeding' and 
`bleeding' of rules, common with transducers (cf.~Section~\ref{r-implementation}).

@@

In practice, we foresee a workflow in which orthography profiles are iteratively
refined, while at the same time inconsistencies and errors in the data to be
tokenized are corrected. In some more complex use-cases there might even be a
need for multiple different orthography profiles to be applied in sequence (see
Sections~\ref{python-implementations} \& \ref{r-implementation} on various exemplary use-cases). The result of any such
workflow will normally be a cleaned dataset and an explicit description of the
orthographic structure in the form of an orthography profile. Subsequently, the
orthography profiles can be easily distributed in scholarly channels alongside
the cleaned data, for example in supplementary material added to journal papers
or in electronic archives.

@@

	\item \textsc{A profile is a delimited text file with an obligatory header
       line}. A minimal profile must have a single column with the header ``Grapheme''. 
	   For any additional columns, the name in the header 
       must be specified. The actual ordering of the columns is unimportant. 
	   The header list must be delimited 
	   in the same way as the rest of the file's contents. Each record must be kept 
	   on a separate line. Separate lines with comments are not allowed. Comments that
       belong to specific lines must be put in a separate column of
       the file, e.g.~add a column called \textsc{comments}.

@@

	\item \textsc{Metadata should be added in a separate UTF-8 text file} using 
	   the JSON-LD dialect specified in \textit{Metadata Vocabulary for Tabular 
	   Data}.\footnote{\url{https://www.w3.org/TR/tabular-metadata/}} This metadata 
	   format allows easy inclusion of Dublin Core metadata,\footnote{\url{http://dublincore.org/}} 
	   which should be used to specify information about the orthographic description 
	   in the orthography profile.\footnote{\url{http://w3c.github.io/csvw/metadata/#dfn-common-property}} 
	   The orthography profile metadata should minimally include provenance information including: 
	   (i) author, (ii) date, (iii) title of the profile, and (iv)
       bibliographic data for resource(s) that illustrate the orthography
       described in the profile. Crucially, the metadata should also specify 
	   (iv) a stable language identifier of the target language of the profile
       using BCP 47/ISO 639-3 or Glottocode as per the CLDF ontology.\footnote{\url{http://cldf.clld.org/v1.0/terms.rdf}}	   	   
	   Further, the metadata file should provide information about the orthography profile's structure and contents, 
	   including: (vi) its dialect description,\footnote{\url{http://w3c.github.io/csvw/metadata/#dfn-dialect-descriptions}} 
	   (vii) the tokenization method (see below), and (ix) proper column           
	   descriptions,\footnote{\url{http://w3c.github.io/csvw/metadata/#dfn-datatype-description}} 
	   which describe how a column should be interpreted and processed (e.g.\ whether they 
	   should be processed as regular expressions; see below).
	   Finally, in accordance with the \textit{Metadata Vocabulary for Tabular 
	   Data}, the metadata's filename should consist of the orthography 
	   profile's filename appended with ``-metadata.json''.\footnote{JSON-LD metadata 
	   is also the choice for datasets conforming to the Cross-Linguistic Data Formats standard, 
	   see: \url{http://cldf.clld.org/}.}

@@

\begin{enumerate}
	\def\labelenumi{A\arabic{enumi}.} \setcounter{enumi}{4} 
	\item \textsc{A minimal profile consists of a single column} with a header
       called \texttt{Grapheme}, listing each of the different graphemes in a
       separate line. The name of this column is crucial for automatic 
       processing.
	\item \textsc{Optional columns can be used to specify the left and right
       context of the grapheme}, to be designated with the headers \texttt{Left}
       and \texttt{Right}, respectively. The same grapheme can occur multiple
       times with different contextual specifications, for example to
       distinguish different pronunciations depending on the context. 
	\item \textsc{The columns \texttt{Grapheme}, \texttt{Left} and \texttt{Right}
       can use regular expression metacharacters.} If regular expressions are
       used, then the must be specified in the metadata file as such, 
	   and all literal usage of the special symbols, like full stops <.>
       or dollar signs <\$> (so-called \textsc{metacharacters}) have to be
       explicitly escaped by adding a backslash before them (i.e.~use
       <\textbackslash.> or <\textbackslash\$>). Note that any specification of
       context automatically expects regular expressions, so it is probably
       better to always escape all regular expression metacharacters when used
       literally in the orthography. The following symbols will need to be
       preceded by a backslash: {[} {]} ( ) \{ \} | ~+ * . - ! ? \^{} \$ and the
       backslash \textbackslash~itself. 
	\item \textsc{An optional column can be used to specify classes of graphemes},
       to be identified by the header \texttt{Class}. For example, this column
       can be used to define a class of vowels. Users can simply add ad-hoc
       identifiers in this column to indicate a group of graphemes, which can
       then be used in the description of the graphemes or the context. The
       identifiers should of course be chosen such that they do not conflate
       with any symbols used in the orthography themselves. Note that such
       classes only refer to the graphemes, not to the context. 
	\item \textsc{Columns describing transliterations for each graphemes can be
       added and named at will}. Often more than a single possible
       transliteration will be of interest. Any software application using these
       profiles should prompt the user to name any of these columns to select a
       specific transliteration. 
	\item \textsc{Any other columns can be added freely, but will mostly be ignored
       by any software application using the profiles}. As orthography profiles
       are also intended to be read and interpreted by humans, it is often
       highly useful to add extra information on the graphemes in further
       columns, like for example Unicode codepoints, Unicode names, frequency of
       occurrence, examples of occurrence, explanation of the contextual
       restrictions, or comments. 
 \end{enumerate}

@@

\begin{enumerate}
	\def\labelenumi{B\arabic{enumi}.} 
	\item \textsc{Each line of a profile will be interpreted as a regular
       expression. }Software applications using profiles can also offer to
       interpret a profile in the literal sense to avoid the necessity for the
       user to escape regular expressions metacharacters in the profile.
       However, this only is possible when no contexts or classes are described,
       so this seems only useful in the most basic orthographies. 
	\item \textsc{The \textsc{class} column will be used to produce explicit
       \textsc{or} chains of regular expressions}, which will then be inserted
       in the \texttt{Grapheme}, \texttt{Left} and \texttt{Right} columns at
       the position indicated by the class-identifiers. For example, a class
       called \texttt{V} as a context specification might be replaced by a regular
       expression like:
       \texttt{(a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au)}.
       Only the graphemes themselves are included here, not any contexts
       specified for the elements of the class. Note that in some cases the 
       ordering inside this regular expression might be crucial, therefore it must be 
	   specified as so in the metadata file.
	\item \textsc{The \textsc{left} and \textsc{right} contexts will be included
       into the regular expressions by using lookbehind and lookahead}.
       Basically, the actual regular expression syntax of lookbehind and
       lookahead is simply hidden to the users by allowing them to only specify
       the contexts themselves. Internally, the contexts in the columns
       \texttt{Left} and \texttt{Right} are combined with the column
       \texttt{Grapheme} to form a complex regular expression like:\ 
       \texttt{(?\textless{}=Left)Grapheme(?=Right)}. 
	\item \textsc{The regular expressions will be applied in the order as specified
       in the profile, from top to bottom.} A software implementation can offer
       help in figuring out the optimal ordering of the regular expressions, but
       then it should be made explicit in the orthograpy profile because regular 
	   expressions are executed in order from top to bottom.
\end{enumerate}

@@

\begin{enumerate}
	\def\labelenumi{B\arabic{enumi}.} \setcounter{enumi}{4} 
	\item \textsc{All graphemes are matched in the text before they are tokenized
       or transliterated}. In this way, there is no necessity for the user to
       consider \textit{feeding} and \textit{bleeding} situations, in which the application of
       a rule either changes the text so another rule suddenly applies (feeding)
       or prevents another rule to apply (bleeding). 
	\item \textsc{The matching of the graphemes can occur either globally or
       linearly.} From a computer science perspective, the most natural way to
       match graphemes from a profile in some text is by walking linearly
       through the text-string from left to right, and at each position go
       through all graphemes in the profile to see which one matches, then go to
       the position at the end of the matched grapheme and start over. This is
       basically how a finite state transducer works, which is a
       well-established technique in computer science. However, from a
       linguistic point of view, our experience is that most linguists find it
       more natural to think from a global perspective. In this approach, the
       first grapheme in the profile is matched everywhere in the text-string
       first, before moving to the next grapheme in the profile. Theoretically,
       these approaches will lead to different results, though in practice of
       actual natural language orthographies they almost always lead to the same
       result. Still, we suggest that any software application using orthography
       profiles should offer both approaches (i.e. \textsc{global} or
       \textsc{linear}) to the user. The approach used should be documented in
       the metadata as \textsc{tokenization method}. 
	\item \textsc{The matching of the graphemes can occur either in NFC or NFD.} By
       default, both the profile and the text-string to be tokenized should be
       treated as NFC (see Section~\ref{pitfall-canonical-equivalence}). However, in some use-cases it turns out to
       be practical to treat both text and profile as NFD. This typically
       happens when very many different combinations of diacritics occur in the
       data. An NFD-profile can then be used to first check which individual
       diacritics are used, before turning to the more cumbersome inspection of
       all combinations. We suggest that any software application using
       orthography profiles should offer both approaches (i.e. \textsc{NFC} or
       \textsc{NFD}) to the user. The approach used should be documented in the
       metadata as \textsc{unicode normalization}. 
	\item \textsc{The text-string is always returned in tokenized form} by
       separating the matched graphemes by a user-specified symbols-string. Any
       transliteration will be returned on top of the tokenization. 
	\item \textsc{Leftover characters}, i.e.~\textsc{characters that are not
       matched by the profile, should be reported to the user as errors.}
       Typically, the unmatched character are replaced in the tokenization by a
       user-specified symbol-string.       
 \end{enumerate}
